Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-11-21 index.qmd
section:issues; process:write

<!-- 2. Thurstone's Theory of CJ: discriminal difference -->
However, since the discriminal process of a single stimulus is not directly observable, the *law of comparative judgment* becomes essential. This law states that in pairwise comparisons, the stimulus positioned further along the trait continuum is perceived as having a higher level of that trait. Thus, the theory assumes the observed dichotomous outcome is determined by the distribution of the difference between the underlying discriminal processes of the stimuli, referred to as the *discriminal difference*. This suggests that the outcome depends on the relative distance between stimuli, rather than their absolute positions on the trait continuum.

<!-- 3. Example-->
These concepts can be illustrated through an example. For instance, in the context of evaluating text quality, @fig-discriminal_process could depict the underlying discriminal process distributions for two written texts, highlighting differences in their discriminal dispersions and modal discriminal processes along the quality trait continuum. Furthermore, @fig-discriminal_difference could display the discriminal difference distribution for these texts, showing that text A is perceived to exhibit significantly higher quality than text B, as indicated by the shaded gray area. Consequently, the dichotomous outcome of this comparison would favor text A.

@@

<!-- 3. What does the zero correlation between stimuli imply? and does it hold? -->
The assumption of *zero correlation between stimuli* is more easily understood through an example. For instance, when using pairwise comparisons to evaluate text quality, the assumption suggests that a judge's perception of the trait in one text does not influence his perception of the same trait on another. Thurstone attributed this independence to the cancellation of potential judges' biases, driven by two opposing and equally weighted effects occurring during the pairwise comparisons [@Thurstone_1927b, pp. 268]. This cancellation was mathematically demonstrated by @Andrich_1978, assuming discriminal processes with additive biases. However, it is easy to imagine two scenarios where the zero correlation assumption almost certainly does not hold: when the pairwise comparison involves multidimensional, complex traits with heterogeneous stimuli, and when an additional hierarchical structure is relevant to the stimuli.

@@

<!-- 4. The first scenario where the assumption does not hold -->
In the first scenario, evidence suggests that when judges are confronted with multi-dimensional, complex traits and heterogeneous stimuli, such as when evaluating written text quality, their assessments are influenced by multiple, intricate aspects of the stimuli [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. In such cases, it is not inconceivable that these various aspects of the stimuli unevenly influence the judges’ perception of one text over another, leading to biases that resist cancellation, as the effects of the factors involved in the comparison are neither equally weighted nor opposing. For example, this may occur when a judge, while assessing the argumentative quality of a text, places disproportionate emphasis on grammatical errors, ultimately favoring texts with fewer errors but weaker arguments. Although the literature does not provide direct evidence for this specific argument, studies such as @Pollitt_et_al_2003 do offer evidence of judges' biases, supporting the idea that the factors involved in pairwise comparisons may not always cancel out.

@@

<!-- 5. The second scenario where the assumption does not hold -->
<!-- In the second scenario, the literature has not even considered the issues related with the existence of an additional hierarchical structure relevant to the stimuli, such as in cases where data includes multiple samples of stimuli within different individuals. In this scenario, there is an amount of correlation that it is not accounted for -->

<!-- 6. But what are the problems when the zero correlation assumption does not hold? -->
<!-- The best the literature has done is to estimate scores and then use those scores to run a multilevel model as in @Boonen_et_al_2020, which in itself brings its own problems. Refer to @sec-theory-issue2 -->

<!-- Such differences may not be detected through analyses of bias and misfit [@Kelly_et_al_2022, pp. 683].  -->

<!-- @Pollitt_et_al_2003 In Thurstone’s method each judge’s standard cancels out when a comparison is made, and any misfit between data and model must indicate a difference of some other kind. This is studied by analyzing the residuals that remain after the model is fitted to the data. Whenever one script is judged to have a higher value than a competing one it is expected to ‘win’ the comparison: if it does the residual will be small, but if it loses then the residual for that judge’s comparison of those two scripts will be large. Any patterns of large residuals will indicate some sort of bias – in this context bias simply means a source of systematic, rather than random, variability. indicates that one judge – Judge 5 – is significantly biased in some way. (Consider the problem of using adhoc methods) -->

@@

<!-- (This a is a validity contention) -->
<!-- Furthermore, it is not inconceivable that the selection of a particular group of judges used to perform the pairwise comparison could result in an incomplete depiction of the dimensions and complexity of the trait. Previous research has highlighted that factors such as age, culture, and education [@Kelly_et_al_2022, pp. 683], as well as individual differences among judges [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020], could influence judgment accuracy, thereby affecting the "shared consensus" of the trait measurement. As @Kelly_et_al_2022 noted [pp. 683], “Would the aggregate view of young, British-Asian men always be the same as the aggregate view of older, black women?” -->

@@

<!-- 4. The problems of assuming equal dispersion between stimuli -->

<!-- it is not hard to imagine that certain individuals can produce good quality texts with more precision than other individuals. Example of exaggerated comparison of individuals, university students vs secondary school students. -->

<!--   here we face misspecification in the model -->

<!-- "It is probably true that this variability of the discriminal dispersion on the psychological continuum is of relatively less serious importance in dealing with strictly homogeneous stimulus series, but it becomes a serious factor in dealing with less conspicuous attributes or with less homogeneous stimulus series such as handwriting specimens, English compositions, sewing samples, oriental rugs." [@Thurstone_1927a, pp. 374]. -->

<!-- Of these, the most questionable is the one that the discriminal dispersions of all the objects are equal. Thurstone clearly did not expect this to hold for any but the most simple stimuli. The scripts used in cross-moderation exercises are obviously far more complex than any used by Thurstone, so it seems naïve to expect this assumption to hold here [@Bramley_2008]. -->

@@

<!-- 3. But what are the challenges in using BTL scores for further analysis? -->
However, the statistical literature advises caution when using estimated scores to conduct additional analyses and tests while warning against eliminating data through ad-hoc univariate procedures. A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty. Ignoring this uncertainty can bias the analysis and reduce the precision of hypothesis tests. Notably, the direction and magnitude of such biases are often unpredictable. Results may be attenuated, exaggerated, or remain unaffected depending on the degree of uncertainty in the scores and the actual effects being tested [@Kline_et_al_2023, pp. 25; @Hoyle_et_al_2023, pp. 137]. Moreover, eliminating data using ad-hoc univariate procedures using these scores compounds this issue by excluding possible valuable information from the analyses [@McElreath_2020], further introducing bias. Finally, the reduced precision in hypothesis tests weakens their statistical power, increasing the likelihood of type-I or type-II errors [@McElreath_2020].

@@

## Assessment design features and their role on reliability and validity {#sec-theory-issue3}

>>>>

<!-- 2. Thurstone's Theory of CJ: discriminal difference -->
However, since the discriminal process of a single stimulus is not directly observable, the *law of comparative judgment* becomes essential. This law states that in pairwise comparisons, the stimulus positioned further along the trait continuum is perceived as having a higher level of that trait. Thus, the theory assumes the observed dichotomous outcome is determined by the distribution of the difference between the underlying discriminal processes of the stimuli, referred to as the *discriminal difference*. This indicates that the outcome depends on the relative distance between stimuli, rather than their absolute positions on the trait continuum.

<!-- 3. Example-->
These concepts are more easily understood through an example. For instance, in the context of evaluating text quality, @fig-discriminal_process could depict the underlying discriminal process distributions for two written texts, highlighting differences in their discriminal dispersions and modal discriminal processes along the quality trait continuum. Furthermore, @fig-discriminal_difference could display the discriminal difference distribution for these texts, showing that text A is perceived to exhibit significantly higher quality than text B, as indicated by the shaded gray area. Consequently, the dichotomous outcome of this comparison would favor text A.

@@

<!-- 3. What does the zero correlation between stimuli imply? and does it hold? -->
The assumption of *zero correlation between stimuli* is more easily understood through an example. For instance, when using pairwise comparisons to evaluate text quality, the assumption suggests that the judge's perception of the trait in one text does not influence his perception of the same trait on another. Thurstone attributed this independence to the cancellation of potential judges' biases, driven by two opposing and equally weighted effects occurring during the pairwise comparisons [@Thurstone_1927b, pp. 268]. This cancellation was mathematically demonstrated by @Andrich_1978, assuming discriminal processes with additive biases. However, it is easy to imagine two scenarios where the zero correlation assumption almost certainly does not hold: when the pairwise comparison involves multidimensional, complex traits with heterogeneous stimuli, and when an additional hierarchical structure is relevant to the stimuli.

@@

<!-- 4. The first scenario where the assumption does not hold -->
In the first scenario, evidence suggests that when judges are confronted with multi-dimensional, complex traits and heterogeneous stimuli, such as when evaluating written text quality, their assessments are influenced by multiple, intricate aspects of the stimuli [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. In such cases, it is not inconceivable that various aspects of the stimuli unevenly influence the judges’ perception of one text over another, leading to biases that resist cancellation, as the effects of the factors involved in the comparison are neither equally weighted nor opposing. For example, this may occur when a judge, while assessing the argumentative quality of a text, places disproportionate emphasis on grammatical errors, ultimately favoring texts with fewer errors but weaker arguments. Although the literature does not provide direct evidence for this specific argument, studies such as @Pollitt_et_al_2003 do offer evidence of judges' biases, supporting the idea that the factors involved in the pairwise comparisons may not always cancel out.

@@

<!-- 5. The second scenario where the assumption does not hold -->
In the second scenario, although the CJ literature has acknowledged the existence of additional hierarchical structures relevant to the stimuli, the statistical handling of this extra source of correlation has been inadequate. For example, in cases where data included multiple samples of stimuli within different individuals, researchers have often relied on using (averaged) estimated BTL scores to conduct further analyses and tests at the hierarchical level of the individuals [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021]. However, this approach has the notable limitation of ignoring the uncertainty associated with the scores (refer to section @sec-theory-issue2 for a detailed discussion of this issue).

<!-- (Is this a validity contention?) -->
<!-- Furthermore, it is not inconceivable that the selection of a particular group of judges used to perform the pairwise comparison could result in an incomplete depiction of the dimensions and complexity of the trait. Previous research has highlighted that factors such as age, culture, and education [@Kelly_et_al_2022, pp. 683], as well as individual differences among judges [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020], could influence judgment accuracy, thereby affecting the "shared consensus" of the trait measurement. As @Kelly_et_al_2022 noted [pp. 683], “Would the aggregate view of young, British-Asian men always be the same as the aggregate view of older, black women?” -->

<!-- 6. What does the equal dispersion between stimuli imply? and does it hold? -->
In contrast, the assumption of *equal dispersion between stimuli* suggests that the perceived trait of all texts varies around their modal discriminal process in the same way across comparisons. While Thurstone noted that the assumption of equal dispersion may be violated when "dealing with less conspicuous attributes or with less homogeneous stimuli" [@Thurstone_1927a, pp. 374], no study explicitly proposes that this assumption may also be violated due to the presence of an additional hierarchical structure relevant to the texts. One such scenario could occur when comparing university and secondary school students. In this case, university students may consistently (or more precisely) produce higher-quality texts, while secondary school students might exhibit a broader range of writing abilities, leading to greater variability in their text quality. Despite the contrived nature of the example, it helps to effectively illustrate how assuming equal dispersions in the texts can overlook meaningful differences in the reliability of text quality across groups or individuals.

<!-- Of these, the most questionable is the one that the discriminal dispersions of all the objects are equal. Thurstone clearly did not expect this to hold for any but the most simple stimuli. The scripts used in cross-moderation exercises are obviously far more complex than any used by Thurstone, so it seems naive to expect this assumption to hold here [@Bramley_2008]. -->

<!-- X. But what are the problems when the zero correlation assumption does not hold? -->

@@

<!-- Different dispersion: here we face misspecification in the model? -->

<!-- This observation highlights how variability in the quality of stimuli should be accounted for when comparing heterogeneous groups or evaluating complex attributes like writing ability. -->

@@

@@

<!-- 3. But what are the challenges in using BTL scores for further analysis? -->
However, the statistical literature advises caution when using estimated scores for additional analyses and tests, as well as when eliminating data through ad hoc univariate procedures. A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty. Ignoring this uncertainty can bias the analysis and reduce the precision of hypothesis tests. Notably, the direction and magnitude of such biases are often unpredictable. Results may be attenuated, exaggerated, or remain unaffected depending on the degree of uncertainty in the scores and the actual effects being tested [@Kline_et_al_2023, pp. 25; @Hoyle_et_al_2023, pp. 137]. Moreover, excluding data using ad hoc univariate procedures can compound these issues by discarding potentially valuable information, further exacerbating the bias [@Zimmerman_1994; @McElreath_2020]. Finally, the reduced precision in hypothesis tests diminishes their statistical power, increasing the likelihood of type-I or type-II errors [@McElreath_2020].

@@

## The assessment design features and their role on reliability and validity {#sec-theory-issue3}
